---
title: "Classic Test Theory"
---


```{r}
library(tidyverse)
library(ggplot2)
```


## Simulating data following CTT

We will simulate IQ test scores from 5000 participants, pretending we know each person's true score.


```{r}
set.seed(777)

# true score parameters
m <- 100 #  mean of true score in the population
s <- 15  # standard deviation of true score in the population

# error score parameters
s_err <- 10

# simulating data
n <- 5000
true_scores <- rnorm(n, m, s)
error_scores <- rnorm(n, 0, s_err)
observed_scores <- true_scores + error_scores


# group into a dataframe
dd <- data.frame(true_scores, error_scores, observed_scores)

```

This is what the data looks like:


```{r}

dd |> 
  pivot_longer(cols = true_scores:observed_scores) |> 
  mutate(name = str_replace_all(name, "_", " ")) |> 
  ggplot(aes(value, fill=name)) + 
  geom_histogram(bins = 50, color="white") + 
  facet_grid(name~.) +
  guides(fill=FALSE)


```


## Exercise: compute reliability index

As a reminder:

$\rho^2_{XT} = \frac{\sigma^2_T}{\sigma^2_X}$

```{r}
# TODO: compute the reliability of the IQ test

```


What's the reliability of our IQ test?



## Focus

Let's focus now on a subset of our participants. For example, let's say we are testing the IQ of the smart students at our University.  

```{r}
dds <- dd |> 
  filter(true_scores > 120)
```

Look at the data:

```{r}

dds |> 
  pivot_longer(cols = true_scores:observed_scores) |> 
  mutate(name = str_replace_all(name, "_", " ")) |> 
  ggplot(aes(value, fill=name)) + 
  geom_histogram(bins = 50, color="white") + 
  facet_grid(name~.) +
  guides(fill=FALSE)


```


What's the reliability of our IQ test now?


```{r}
# TODO: compute the reliability of the IQ test

```


What to you conclude from this? 


