---
title: "Classic Test Theory"
---


```{r}
library(tidyverse)
```


## Simulating data following CTT

We will simulate IQ test scores from 5000 participants, pretending we know each person's true score.


```{r}
set.seed(777)

# true score parameters
m <- 100 #  mean of true score in the population
s <- 15  # standard deviation of true score in the population

# error score parameters
s_err <- 10

# simulating data
n <- 5000
true_scores <- rnorm(n, m, s)
error_scores <- rnorm(n, 0, s_err)
observed_scores <- true_scores + error_scores


# group into a dataframe
dd <- data.frame(true_scores, error_scores, observed_scores)

```

This is what the data looks like:


```{r}

dd |> 
  pivot_longer(cols = true_scores:observed_scores) |> 
  mutate(name = str_replace_all(name, "_", " ")) |> 
  ggplot(aes(value, fill=name)) + 
  geom_histogram(bins = 50, color="white") + 
  facet_grid(name~.) +
  guides(fill=FALSE)


```


## Exercise: compute reliability index

As a reminder:

$\rho^2_{XT} = \frac{\sigma^2_T}{\sigma^2_X}$

```{r}
# TODO: compute the reliability of the IQ test

```


What's the reliability of our IQ test?



## Focus

Let's focus now on a subset of our participants. For example, let's say we are testing the IQ of the smart students at our University.  

```{r}
dds <- dd |> 
  filter(true_scores > 120)
```

Look at the data:

```{r}

dds |> 
  pivot_longer(cols = true_scores:observed_scores) |> 
  mutate(name = str_replace_all(name, "_", " ")) |> 
  ggplot(aes(value, fill=name)) + 
  geom_histogram(bins = 50, color="white") + 
  facet_grid(name~.) +
  guides(fill=FALSE)


```


What's the reliability of our IQ test now?


```{r}
# TODO: compute the reliability of the IQ test

```


What do you conclude from this? 





## Repeated measures

Add a repetition to the simulated data

```{r}
# TODO: simulate the scenario where the same test is completed twice by the same people.


```


Compute the reliability using the repeated tests

```{r}
# TODO: compute the reliability of the IQ test using the repeated tests.

```


Do you get the same value as when using the previous formula (that uses true scores)?



## Our own data

Here are the results from our little study. The same "test" was completed 3 times. 

```{r}
data <- data.frame(
  Q1 = c(5, 4, 2, 5, 7, 7, 7, 6, 7, 4, 3),
  Q2 = c(5, 4, 2, 5, 5, 7, 6, 6, 7, 4, 3),
  Q3 = c(5, 5, 4, 6, 5, 7, 8, 7, 8, 4, 4)
)
```



Compute the reliability of this test. Does the reliability change across time?


```{r}


```



## Standard error of the mean

Going back to the IQ example, assume we have a participant that has an observed IQ score of 120. What is the standard error of the mean for this IQ test? What is the 95% confidence interval for this IQ score of 120?



```{r}
# TODO: compute 95% CI for observed IQ score of 120

```





## Questions

Do you think the reliabilty index as defined by CTT is satisfactory?

> Enter your response here


If your patient fills out a questionnaire and gets a score of 19, what wouly you like a reliability index to tell you about this number?

> Enter your response here


